The role of aneuploidy in the evolution of cancer drug resistance

### 2023-07-24 EvolutionaryRescue1.tex
fixed tau1 errors; removed linear approx for tau1 and tau2; added main messages and figures outline in comments

@@

\paragraph{Evolutionary rescue time.}
In Appendix C we derive an approximation for $\tau_1$, the expected rescue time without aneuploidy ($u=0$), and $\tau_2$, the expected rescue time with aneuploidy ($u>0$),
\begin{align}
\tau_1 &\approx \frac{1+\e^{-Np_w}}{u N p_a} , \label{eq:limitapprox} ,\\
\tau_2 &\approx \frac{1+\e^{-Np_w}}{u N p_m} , \label{eq:limitapprox3} .
\end{align}
\Cref{MeanTimeGrowthAneuploidyPlot} shows the agreement between this approximations and simulation results for intermediate and large tumor sizes.

Thus, the ratio of the expected rescue time with and without aneuploidy ($u>0$ and $u=0$, respectively),
\begin{align}\label{eq:eqMeanTimeRatioInitialPopulationSize}
\frac{\tau_1}{\tau_2}  \approx \frac{p_a}{p_m} = 
%&\frac{\lambda_m}{\lambda_a} \; \frac{\Delta_a-v+\sqrt{(\Delta_a - v)^2 + (1/T^*)^2}}{2 \Delta_m} \approx \\
%&\frac{1}{2} \; \frac{\lambda_m}{\Delta_m} \left[ \left(\frac{\lambda_a}{\Delta_a}\right)^{-1} + \sqrt{\frac{\left(\Delta_a\right)^2 + (1/T^*)^2}{\lambda_a^2}}\right] = \\
\frac{1}{2} \; \frac{\lambda_m}{\Delta_m} \left[ \left(\frac{\lambda_a}{\Delta_a}\right)^{-1} + \sqrt{\left(\frac{\lambda_a}{\Delta_a}\right)^{-2} + 4v \left(\lambda_a\frac{\lambda_m}{\Delta_m}\right)^{-1}}\right] .
\end{align}

\Cref{MeanTimeRatioInitialPopulationSize} shows how this ratio, $\tau_1/\tau_2$, changes as a function of the initial tumor size, $N$, for different aneuploid growth rates, $\Delta_a$.

Importantly, in \Cref{MeanTimeGrowthAneuploidyPlot} we observe that in initially large tumors, direct mutation drives evolutionary rescue while aneuploidy plays a role for intermediate sized tumors. % TODO where do you see this..?
This is consistent with \Cref{rescue_ratio}B,D, where aneuploidy only increases the probability of evolutionary rescue in small and intermediate tumors ($N<10^7$).

@@

@@

\section*{Appendix C: Evolutionary rescue time}

@@

The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = u p_a w_t$,
where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
Note that 
\begin{equation}\label{eq:integralR}
\int_0^{t}{R(z)\d z} = 
u p_a N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
u p_a N t,
\end{equation}
by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w}=t+O(t^2)$.
The probability density function of $T_1$ is thus
$R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)$. % TODO ref
Therefore, the probability density function of the conditional random variable $(T_1 \mid T_1 < \infty)$ is
$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{p_{rescue}}$. 
\\

We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
\begin{equation}\label{eq:meantime1}
\begin{aligned}
\tau_1 =
\int_{0}^{\infty}{t f_1(t) \d t} = 
\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}} = 
\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}}
\end{aligned}
\end{equation}
after applying integration by parts.
Therefore, plugging \cref{eq:integralR,eq:rescue_prob} in \cref{eq:meantime1}, 
\begin{align}\label{eq:limitapprox_appendix}
\tau_1 = 
%\frac{\int_0^\infty\exp\left(-\int_0^\tau R(z)\d z\right)\d t}{1-\left(1-p_w\right)^N} = 
\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w t}-1}{\Delta_w}} \d t}{1-\left(1-p_w\right)^N} \approx
\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx \\
\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u p_a N t} \d t =
\frac{1+\e^{-Np_w}}{u p_a N} ,
\end{align}
where we use the approximations 
$\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2)$ and $(1-\e^{-Np_w})^{-1} \approx 1+\e^{-Np_w}$ and integrate the exponent.
\Cref{MeanTimeGrowthAneuploidyPlot}B show the agreement between this approximating and simulation results for intermediate and large tumor sizes.
\\

>>>>

% MESSAGES
% 1. aneuploidy increases the prob of rescue -- reduces the tumor threshold size for rescue
% 2. aneuploidy changes the survival curve in a distinct way

% FIGURES
% Fig 1A: model graph chart
% Fig 1B: Delta_w, Delta_a, Delta_m illustration

% Fig 2: example simulations - w_t, a_t, m_t over time 
% 	A u=0; B Delta_a << 0; C Delta_a ~ 0; D Delta_a >>0

% Fig 3A: prob rescue vs N for various Delta_a; markup N*
% Fig 3B: N* vs Delta_a
% Fig 3C: N* vs u/v

% Fig 4: survival curves
% 	A u=0; B Delta_a << 0; C Delta_a ~ 0; D Delta_a >>0

@@

\paragraph{Evolutionary rescue time.}
In Appendix C we derive an approximation for $\tau_1$, the expected rescue time without aneuploidy ($u=0$), and $\tau_2$, the expected rescue time with aneuploidy ($u>0$). \Cref{MeanTimeGrowthAneuploidyPlot} shows the agreement between these approximations and simulation results for intermediate and large tumor sizes.

@@

% ? Distribution of recurrence time -- with and without aneuploidy

% Kaplan-Meier curves -- x: time; y: P(# cells >= N) -- once for u=0; u>0 with the three scenarios

@@

\section*{Appendix C: Evolutionary rescue time}
% TODO remove the linear approximation, it is not good.

@@

The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = v p_m w_t$,
where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
Note that 
\begin{equation}\label{eq:integralR}
\int_0^{t}{R(z)\d z} = 
v p_m N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
v p_m N t,
\end{equation}
by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w}=t+O(t^2)$.
The probability density function of $T_1$ is thus
$R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)$. % TODO ref
Therefore, the probability density function of the conditional random variable $(T_1 \mid T_1 < \infty)$ is
$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{\presc}$. 
\\

We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
\begin{equation}\label{eq:meantime1}
\begin{aligned}
\tau_1 =
\int_{0}^{\infty}{t f_1(t) \d t} = 
\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}} = 
\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}}
\end{aligned}
\end{equation}
after applying integration by parts.
Therefore, plugging \cref{eq:integralR,eq:rescue_prob} in \cref{eq:meantime1}, 
\begin{align}\label{eq:limitapprox_appendix}
\tau_1 = 
%\frac{\int_0^\infty\exp\left(-\int_0^\tau R(z)\d z\right)\d t}{1-\left(1-p_w\right)^N} = 
\frac{\int_0^\infty\e^{-v N p_m\frac{\e^{\Delta_w t}-1}{\Delta_w}} \d t}{1-\left(1-p_w\right)^N} \approx
\frac{\int_{0}^{\infty}{\exp\left(-v p_m N t\right) \d t}}{1-\e^{-N p_w}} \approx \\
\left(1+\e^{-N p_w}\right)\int_0^\infty\e^{-v p_m N t} \d t =
\frac{1+\e^{-N p_w}}{v p_m N} ,
\end{align}
where we use the approximations 
$\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2)$ and $(1-\e^{-N p_w})^{-1} \approx 1+\e^{-N p_w}$ and integrate the exponent.
\Cref{MeanTimeGrowthAneuploidyPlot}B show the agreement between this approximating and simulation results for intermediate and large tumor sizes.
\\
